The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2012-01-25 smurf_paper.tex
added in an old reference that might be relevant
more plots for cosmology section
flesh out text for cosmology section

@@

\subsection{Point source}

The accurate measurement of positions and brightnesses of point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomical units.  In this example we produce a 450\,\micron\ map
of Uranus (observation 26 on 20111017), which is a nearly point-like
source for SCUBA-2 that is commonly used as a primary flux
calibrator. The now standard CV daisy scan pattern was used, with a
scan speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of different model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the S/N of Uranus is clearly
large ($\sim$175 as estimated by SMURF), enabling us to see the faint
sidelobes (circle and cross pattern), the map also has obvious
circular streaks and other large-scale ripples. The circular streaks
are due to the fact that \model{COM} does not account for all of the
low-frequency noise (see Fig.~\ref{fig:pspec}), and therefore each
bolometer leaves a trace of the circular scan pattern in the map as
each of their baselines slowly drift independently. Most of the
larger-scale ripples in the map, however, are probably due to
degeneracies in the map solution.

To illustrate how large-scale ripples can form (and in fact, diverge),
the same map solution is run for 50 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 50 (grey) iterations which are nearly
identical, yet the \emph{change} in the estimated \model{COM} (green)
and \model{AST} (red) signals between 2 and 50 iterations are
large. However, it is also clear that the estimated \model{COM} and
\model{AST} signals are \emph{complimentary}. In other words, the large
change in the \model{AST} signal is cancelled by freedom in the
\model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the pointing solution
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

One simple improvement is to simply restrict the low-frequency signal
that goes into the maps using a high-pass filter. A third reduction of
the data uses the default map-making parameters, as described in
Section~\ref{sec:algorithm}, which adds the \model{FLT} model to
accomplish this task immediately prior to map estimation. We set the
filter edge based on an angular scale of 200\,arcsec, which given the
scan speed of 155\,arcsec\,sec$^{-1}$, corresponds to a frequency of
0.78\,Hz. Now using the automatic map-based convergence test (Section
???), the resulting map converges after 10 iterations
(Fig.~\ref{fig:pointmaps}c). Both the circular streaks and the
large-scale gradient are now removed, but they have been replaced by
an obvious, circularly-symmetric ringing pattern about Uranus. The
reason for this ringing is due to the fact that the hard-edged
high-pass filter in frequency space is equivalent to a $\sinc$
function-like response in map space. Since the scan pattern is
isotropic (scans equally at all position angles), and there is a
bright point-like source, the result is an azimuthally-symmetric
$\sinc$ function-like pattern in the map about Uranus. Note that due
to the iterative nature of the solution, much of the signal in the
source is actually removed for subsequent iterations of
\model{FLT}, but as in the previous case, generacies between the
map and the \model{FLT} and \model{COM} signal components can still
lead to complicated structures.

The preceding examples illustrate the need for prior constraints in
the map solution. For calibrators (and other previously known bright,
compact sources), a good, simple prior is to constrain the map to a
value of zero away from the known locations of emission. In
Fig.~\ref{fig:pointmaps}d, a map is produced in an identical manner to
the previous example, but now setting the map explicitly to zero
beyond a radius of 60\,arcsec from the location of Uranus (much larger
than the FWHM of the main lobe), for all but the final iteration. In
this case, the map converges after 22 iterations, and all of the
previous artifacts have been effectively removed. Since the map is now
flat away from the source, and constrained to a value of zero, it is
now appropriate for performing aperture photometry directly, with no
need for an additional reference annulus. This approach to constrained
map-making is similar to that employed for poorly cross-linked scans
of compact (though resolved) sources by \citet{wiebe2009} using BLAST
data.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps.pdf}
\caption{Multiple reductions of a 450\,\micron\ daisy scan of Uranus,
  all scaled between -0.002\,pW (white) and +0.002\,pW (black): (a)
  Reduction in which only common-mode subtraction is used to suppress
  low-frequency noise, and the reduction is forced to use 2 iterations
  (after the first iteration an estimate of the source flux is
  removed, and the noise is measured in the residuals to obtain
  appropriate weighting for the second and final iteration) ---
  circular streaks are caused by independent low-frequency noise that
  is not removed by the common-mode (Uranus peak 0.14\,pW); (b) same
  as (a) but using 50 iterations, illustrating that there is a
  degeneracy between large-scale structure and the common-mode (the
  footprint of working bolometers is also shown for reference, Uranus
  peak 0.14\,pW); (c) reduction in which high-pass filtering above
  0.775\,Hz (corresponding to a spatial scale of 200\,arcsec as
  indicated) is applied after common-mode removal, but before the map
  estimate --- the map-based convergence test is activated and the
  solution halts after 10 iterations, but leaving large-scale ringing
  due to the filter (Uranus peak 0.38\,pW); (d) same as (c), but now
  the region of the map beyond the white dot-dashed circle is
  constrained to a value of zero until all but the final iteration ---
  the map is now extremely flat, there is little attenuation of the
  source flux, and the diffraction pattern is clearly seen (Uranus
  peak 0.38\,pW).}
\label{fig:pointmaps}
\end{figure*}

@@

\subsection{Deep blind survey}

\scuba\ surveys designed to detect extremely faint point-sources
(e.g. high-redshift star-forming galaxies, and debris disks) need to
obtain the best noise performance possible. The approach described
here involves three major steps: (i) generating a map that removes
most low-frequency noise sources; (ii) \emph{whitening} (essentially
flattening) the map to make the noise properties of the map easier to
deal with; and (iii) detecting point sources using a
\emph{matched-filter}. Note that variations on this general procedure
have been used extensively in the submm cosmology community using
previous instruments \citep[e.g.][]{scott2008,perera2008}.  In this
section we will reduce data of the Lockman Hole taken in 2010 during
S2SRO as a pilot project for the Cosmology Legacy Survey. It consists
of $\sim8.5$\,hours of data taken in 36 separate scans (average
15\,min. each) spread over February and March, covering an area of
about 50\,arcmin$^2$. ... full obs list ... dropped scans due to
terrible noise properties ...

The first step, map generation, is accomplished in a way similar to
the default described in the previous section (without any boundary
constraints)...

Even though the map \emph{looks} extremely flat, it is clear that
there is a mixture of faint astronomical sources, and what is probably
residual low-frequency noise causing faint ripples. With uncertainties
about what is source, and what is an artifact, the first step is to
remove the low-frequency noise in a way that does not affect
(appreciably) the real sources. To accomplish this goal, the map is
run through a whitening filter that results in a flat angular noise
power spectrum. The noise power spectrum is estimated from a
\emph{jackknife map}: maps are produced from two independent halves of
the total data set, and then a pixel-by-pixel difference is
calculated. Provided that the noise in one half of the data is
uncorrelated with that from the other half, the noise in the jackknife
map should resemble that of the total map, though larger by a factor
of $\sim\sqrt(2)$ depending on how closely the pixel variances in the
two maps are matched (if one map pixel is significantly noisier, it
will dominate the noise in the jackknife).  The astronomical signal,
however, should be cleanly removed (provided that there are no strong
time-varying signals, and also assuming that errors due to calibration
between the two halves are insignificant). The approach we have taken
to minimize systematics is to produce two halves of the jackknife
using odd and even scan numbers (i.e., each map will contain an even
sampling of data across the full dataset).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_rawmaps.pdf}
\caption{Procedure for producing maps of deep fields.}
\label{fig:lockman_rawmaps}
\end{figure}

>>>>

Another paper worth citing? \citet{weferling2002}

@@

\subsection{Known point source}
\label{sec:point}

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomical units.  In this example we produce a 450\,\micron\ map
of Uranus (observation 26 on 20111017), which is a nearly point-like
source for SCUBA-2 that is commonly used as a primary flux
calibrator. The now standard CV daisy scan pattern was used, with a
scan speed of 155 arcsec\,sec$^{-1}$. We perform several different
reductions of the data to illustrate the purpose of different model
components and the convergence properties of the solution
(Fig.~\ref{fig:pointmaps}). In all cases the maps are produced on a
grid of azimuth (horizontal) and elevation (vertical) offsets from the
position of Uranus (the origin), using 2\,arcsec pixels.

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the S/N of Uranus is clearly
large ($\sim$175 as estimated by SMURF), enabling us to see the faint
sidelobes (circle and cross pattern), the map also has obvious
circular streaks and other large-scale ripples. The circular streaks
are due to the fact that \model{COM} does not account for all of the
low-frequency noise (see Fig.~\ref{fig:pspec}), and therefore each
bolometer leaves a trace of the circular scan pattern in the map as
each of their baselines slowly drift independently. Most of the
larger-scale ripples in the map, however, are probably due to
degeneracies in the map solution as we discuss below.

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 50 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 50 (grey) iterations which are nearly
identical, yet the \emph{change} in the estimated \model{COM} (green)
and \model{AST} (red) signals between 2 and 50 iterations are
large. However, it is also clear that the estimated \model{COM} and
\model{AST} signals are \emph{complimentary}. In other words, the
large change in the \model{AST} signal is cancelled by freedom in the
\model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the pointing solution
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

One improvement is to simply restrict the low-frequency signal that
goes into the maps by applying a high-pass filter to the bolometer
time-series. A third reduction of the data uses the default map-making
parameters, as described in Section~\ref{sec:algorithm}, which adds
the \model{FLT} model to accomplish this task immediately prior to map
estimation. We set the filter edge based on an angular scale of
200\,arcsec, which given the scan speed of 155\,arcsec\,sec$^{-1}$,
corresponds to a frequency of 0.78\,Hz. Now using the automatic
map-based convergence test (Section ???), the resulting map converges
after 10 iterations (Fig.~\ref{fig:pointmaps}c). Both the circular
streaks and the large-scale gradient are now removed, but they have
been replaced by an obvious, circularly-symmetric ringing pattern
about Uranus. The reason for this ringing is that the hard-edged
high-pass filter in frequency space is equivalent to a $\sinc$
function-like response in map space. Since the scan pattern is
isotropic (scans equally at all position angles), and there is a
bright point-like source, the result is an azimuthally-symmetric
$\sinc$ function-like pattern in the map about Uranus. Note that due
to the iterative nature of the solution, much of the signal in the
source is actually removed for subsequent iterations of \model{FLT},
but as in the previous case, degeneracies between the map and the
\model{FLT} and \model{COM} signal components can still lead to
complicated structures.

The preceding examples illustrate the need for constraints in the map
solution in many situations. For calibrators (and other previously
known bright, compact sources), a good, simple prior is to constrain
the map to a value of zero away from the known locations of
emission. In Fig.~\ref{fig:pointmaps}d, a map is produced in an
identical manner to the previous example, but now setting the map
explicitly to zero beyond a radius of 60\,arcsec from the location of
Uranus (much larger than the FWHM of the main lobe), for all but the
final iteration. In this case, the map converges after 22 iterations,
and all of the previous artifacts have been effectively removed. Since
the map is now flat away from the source, and constrained to a value
of zero, it is appropriate for performing aperture photometry
directly, with no need for an additional reference annulus. This
approach to constrained map-making is similar to that employed for
poorly cross-linked scans of compact (though resolved) sources by
\citet{wiebe2009} using BLAST data.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps.pdf}
\caption{Multiple reductions of a 450\,\micron\ daisy scan of Uranus,
  all scaled between -0.002\,pW (white) and +0.002\,pW (black): (a)
  Reduction in which only common-mode subtraction is used to suppress
  low-frequency noise, and the reduction is forced to use 2 iterations
  (after the first iteration an estimate of the source flux is
  removed, and the noise is measured in the residuals to obtain
  appropriate weighting for the second and final iteration) ---
  circular streaks are caused by independent low-frequency noise that
  is not removed by the common-mode (Uranus peak 0.14\,pW); (b) same
  as (a) but using 50 iterations, illustrating the degeneracy between
  large-scale structure and the common-mode (the footprint of working
  bolometers is also shown for reference, Uranus peak 0.14\,pW); (c)
  reduction in which high-pass filtering above 0.775\,Hz
  (corresponding to a spatial scale of 200\,arcsec as indicated) is
  applied after common-mode removal, but before the map estimate ---
  the map-based convergence test is activated and the solution halts
  after 10 iterations, but leaving large-scale ringing due to the
  filter (Uranus peak 0.38\,pW); (d) same as (c), but now the region
  of the map beyond the white dot-dashed circle is constrained to a
  value of zero until all but the final iteration --- the map is now
  extremely flat, there is little attenuation of the source flux, and
  the diffraction pattern is clearly seen (Uranus peak 0.38\,pW).}
\label{fig:pointmaps}
\end{figure*}

@@

\subsection{Deep blind survey}
\label{sec:cosmo}

\scuba\ surveys designed to detect extremely faint point-sources
(e.g. high-redshift star-forming galaxies, and debris disks) are
ideally limited by the white-noise performance of the instrument. The
approach described here for maximizing the S/N of point-sources
involves three major steps: (i) generating a map that removes most
low-frequency noise sources with approximately linear response without
prior knowledge of the location of sources; (ii) \emph{whitening}
(essentially flattening) the map to make the noise properties of the
map easier to deal with; and (iii) detecting point sources using a
\emph{matched-filter}. Note that variations on this general procedure
have been used extensively in the submm cosmology community using
previous instruments
\citep[e.g.][]{scott2002,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we will reduce data of the Lockman Hole taken in 2010
during S2SRO as a pilot project for the Cosmology Legacy Survey. It
consists of $\sim8.5$\,hours of data taken in 36 separate scans
(average 15\,min. each) spread over February and March, covering an
area of about 50\,arcmin$^2$. ... full obs list ... dropped scans due
to terrible noise properties ...

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{lockman_maps.pdf}
\caption{Reduction of a blank-field survey: the Lockman Hole. (a) raw
  output of SMURF using high-pass filtering as a pre-processing step,
  followed by 4 iterations using only the \model{COM} model to remove
  residual low-frequency noise. White contours correspond to estimated
  noise levels of 1.25, 2.5 and 5.0 times the minimum noise at the
  centre of the map. (b) the whitened map using a filter based on the
  angular power spectrum of the jackknife map. (c) The whitened map
  cross-correlated with the whitened PSF to identify point
  sources. The image shows the S/N, with 3.8-$\sigma$ peaks indicated
  with blue circles (radius 8\,arcsec). The orange ``$+$'' give the
  locations of radio sources from ... . Of the 10 submm peaks, 9 are
  within this search radius of at least one radio source. (d)
  jackknife map produced from the difference of two maps ,using the
  even and odd scan numbers, respectively. Provided that all noise
  sources are statistically un-correlated between the two halves of
  the data, the map is a plausible realization of the noise without
  contamination from astronomical sources. (e) The jackknife map
  whitened using the same filter as that in panel (b). The whitened
  jackknife map cross-correlated with the same whitened PSF as in
  panel (c). Again, orange ``$+$'' and blue circles indicate radio
  sources and 3.8-$\sigma$ peaks, respectively. Unlike panel (c),
  there is only a single significant peak, and it is not in the
  vicinity of a radio source.}
\label{fig:lockman_maps}
\end{figure*}

The first step, map generation, is different from that described in
Section~\ref{sec:point} in two key ways. Since the locations of
sources are unknown \emph{a priori}, a map constraint is not
employed. Large-scale diverging structures in the map must be
mitigated, and the method used in this example (the default processing
in SMURF) is to apply a high-pass filter to the data \emph{once} as a
pre-processing steps. The iterative solution is then run using only
\model{COM}, \model{EXT}, \model{AST}, and \model{NOI}. In other
words, there is no information in the bolometer signals below some
cutoff frequency, and residual noise is removed only using iterative
estimation of the residual (higher-frequency) common-mode. The map is
shown in Fig.~\ref{fig:lockman_maps}a. The map-maker has been tested
in two ways: (i) large numbers of iterations are used to verify that
the maps converge without the growth of large structure; and (ii)
adding synthetic sources to the data at a range of brightnesses verify
that the map-maker response to them is \emph{linear} (i.e. the
relative shape and amplitude compared to the input source is
independent of brightness). The results of this second test are shown
by the solid and dotted lines in Fig.~\ref{fig:psf},
respectively. Clearly the down-side of using a high-pass filter as a
pre-processing step, and having no other map-constraints, has the
down-side of introducing sidelobes around the main peak. ... high-pass
filter + iterative common-mode removal takes place of PCA cleaning in
AzTEC reduction ... probably more linear.

Even though the map \emph{looks} extremely flat, there is a mixture of
faint astronomical sources, and what is probably residual
low-frequency noise causing faint ripples visible to the naked
eye. Since the mixture of the two components is unknown, the first
step is to remove the low-frequency noise in a way that does not
affect (appreciably) the real sources. To accomplish this goal, the
map is run through a whitening filter that ideally results in a flat
angular noise power spectrum. The noise power spectrum is estimated
from a \emph{jackknife map}: maps are produced from two independent
halves of the total data set, and then a pixel-by-pixel difference is
calculated. Specifically, the jackknife signal in a map pixel,
$S_\mathrm{JK}$, and variance, $\sigma^2_\mathrm{JK}$ are estimated
from the map quantities, $S_1$, $S_2$, $\sigma^2_1$, and $\sigma^2_2$
as,
%
\begin{eqnarray}
S_\mathrm{JK} = \frac{S_1 - S_2}{2} \\
\sigma^2_\mathrm{JK} = \frac{\sigma^2_1 + \sigma^2_2}{2}
\end{eqnarray}

Provided that the noise in one half of the data is uncorrelated with
that from the other half, the noise in the jackknife map should
resemble that of the raw map. The astronomical signal, however, should
be cleanly removed (provided that there are no strong time-varying
signals, and also assuming that errors due to calibration between the
two halves are insignificant). The approach we have taken to minimize
systematics is to produce two halves of the jackknife using odd and
even scan numbers (i.e., each map will contain an even sampling of
data across the full dataset).

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_psf.pdf}
\caption{Slice through the angular response to an ideal Gaussian point
  source (solid line) along the R.A. axis following map-making using
  the default blank-field processing with SMURF (dotted line), and
  upon application of the whitening filter (dashed line). The ``map
  filtered'' response is produced by adding the ideal Gaussian to the
  real data near the centre of the Lockman Hole map, and re-reducing
  the data. The resulting dotted line gives the expected shape of a
  source in panel (a) from Fig.~\ref{fig:lockman_maps}. Applying the
  whitening filter (reciprocal of the solid orange line in
  Fig.~\ref{fig:lockman_pspec}) to the dotted line gives the whitened
  line, which is the expected shape of a source in panel (b) from
  Fig.~\ref{fig:lockman_maps}. Finally, cross-correlating the whitened
  map with this whitened PSF is an effective ``matched filter'' for
  identifying point-like sources, and this smoothed map is shown in
  panel (c) of Fig.~\ref{fig:lockman_maps}.}
\label{fig:lockman_psf}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_pspec.pdf}
\caption{Azimuthally-averaged angular power spectral densities for the
  raw map output by SMURF (dashed black line), the jackknife (solid
  orange line), and whitened (solid black line) maps (panels (a), (d),
  and (b) from Fig.~\ref{fig:lockman_maps}, respectively. Since the
  raw map contains spatially correlated signals on large scales both
  due to noise and astronomical signals (low-spatial frequencies), the
  jackknife map (difference of two approximately equal-length subsets
  of the data) is used to generate a plausible realization of pure
  noise. Assuming that the noise properties are isotropic, a whitening
  filter is estimated from the reciprocal of the jackknife power
  spectrum with only a radial dependence. The power spectrum of the
  resulting signal map has residual power on large scales which is
  presumably due to astronomical sources.}
\label{fig:lockman_pspec}
\end{figure}
